Unshuffling Permutations

### 2016-02-02 new-article.tex
add new-article.tex

@@

@@

@@

@@

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item The permutation $\pi$ is a square.
    \item There exists a perfect containment-free matching $\mathcal{M}$
    of $G_\pi$ such that for any two distinct edges $(a, b)$ and
    $(a', b')$ in $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only
    if $\pi(b) < \pi(b')$.
  \end{enumerate}
\end{lemma}

\begin{proof}[of Lemma~\ref{lemma:matching}]
    The proof is, naturally enough, a slight extension of
    Lemma~1 in \cite{Rizzi:Vialette:CSR:2013} and is added here in full
    for the sake of completeness.

    (\emph{1} $\Rightarrow$ \emph{2}).
    Since $\pi$ is a square, it admits a square root $\sigma$. Then, by
    definition, there are two sets $I^1 = \{i^1_1 < i^1_2 < \dots < i^1_n\}$
    and $I^2 = \{i^2_1 < i^2_2 < \dots < i^2_n\}$ of disjoint indexes of
    letters of $\pi$ such that $\pi_{|I^1}$ and $\pi_{|I^2}$ are both
    order-isomorphic to $\sigma$. Let $2n = |\pi|$ (and hence
    $|\sigma| = n$). Let $G_\pi$ be the linear clique associated to $\pi$
    with vertices
    $\VS{G_\pi} = \{i^1_j : j \in [n]\} \cup \{i^2_j : j \in [n]\}$. It
    is easily seen that $\mathcal{M} = \{\{i^1_j, i^2_j\} : j \in [n]\}$
    is a subset of $\ES{G_\pi}$. Furthermore, $\mathcal{M}$ is a perfect
    matching since $I^1 \cap I^2 = \emptyset$, and
    $I^1 \cup I^2 = \{1, 2,  \dots 2n\}$. It is also containment-free.
    Indeed, if it is not the case then there would exist two edges
    $e = \{i^1_j, i^2_j\}$ and $e' = \{i^1_k, i^2_k\}$, $j < k$, in
    $\mathcal{M}$ such that $i^1_j < i^1_k < i^2_k < i^2_j$ or
    $i^1_j < i^2_k < i^1_k < i^2_j$. This is a contradiction since
    $i^2_k > i^2_j$ if $k > j$. Finally, for any two distinct edges
    $e = \{\i^1_j, i^2_j\}$ and $e' = \{i^1_k, i^2_k\}$ of $\mathcal{M}$,
    we have $\pi(i^1_j) < \pi(i^2_j)$ if and only if
    $\pi(i^1_k) < \pi(i^2_k)$ since we are comparing in both cases two
    elements (at position $j$ and $k$) in two patterns that are
    order-isomorphic to $\sigma$.

>>>>

\newcommand{\LabeledInclusionLL}[4]{%
\begin{tikzpicture}[scale=0.25,inner sep=.8pt,node distance=.45cm,>=latex']
\draw node [] (U1)               {#1};
\draw node [] [right of=U1] (U2) {#2};
\draw node [] [right of=U2] (U3) {#3};
\draw node [] [right of=U3] (U4) {#4};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,1) $) and ($ (U3.north) + (0,1) $) .. (U3.north);
\end{tikzpicture}
}% end \LabeledInclusionLL

@@

\newcommand{\LabeledInclusionLR}[4]{%
\begin{tikzpicture}[scale=0.25,inner sep=.8pt,node distance=.45cm,>=latex']
\draw node [] (U1)               {#1};
\draw node [] [right of=U1] (U2) {#2};
\draw node [] [right of=U2] (U3) {#3};
\draw node [] [right of=U3] (U4) {#4};
\draw [->] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\draw [<-] (U2.north) .. controls ($ (U2.north) + (0,1) $) and ($ (U3.north) + (0,1) $) .. (U3.north);
\end{tikzpicture}
}% end \LabeledInclusionLR

@@

\newcommand{\LabeledInclusionRL}[4]{%
\begin{tikzpicture}[scale=0.25,inner sep=.8pt,node distance=.45cm,>=latex']
\draw node [] (U1)               {#1};
\draw node [] [right of=U1] (U2) {#2};
\draw node [] [right of=U2] (U3) {#3};
\draw node [] [right of=U3] (U4) {#4};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\draw [->] (U2.north) .. controls ($ (U2.north) + (0,1) $) and ($ (U3.north) + (0,1) $) .. (U3.north);
\end{tikzpicture}
}% end \LabeledInclusionRL

@@

\newcommand{\LabeledInclusionRR}[4]{%
\begin{tikzpicture}[scale=0.25,inner sep=.8pt,node distance=.45cm,>=latex']
\draw node [] (U1)               {#1};
\draw node [] [right of=U1] (U2) {#2};
\draw node [] [right of=U2] (U3) {#3};
\draw node [] [right of=U3] (U4) {#4};
\draw [<-] (U1.north) .. controls ($ (U1.north) + (0,2) $) and ($ (U4.north) + (0,2) $) .. (U4.north);
\draw [<-] (U2.north) .. controls ($ (U2.north) + (0,1) $) and ($ (U3.north) + (0,1) $) .. (U3.north);
\end{tikzpicture}
}% end \LabeledInclusionRR

@@

\begin{proof}[of Lemma~\ref{lemma:matching}]
    (\emph{1} $\Rightarrow$ \emph{2}).
    Since $\pi$ is a square, it admits a square root, say $\sigma$.
    Let $2n = |\pi|$ (and hence $|\sigma| = n$).
    Then, by definition, there exist two sets
    $I^1 = \{i^1_1 < i^1_2 < \dots < i^1_n\}$
    and $I^2 = \{i^2_1 < i^2_2 < \dots < i^2_n\}$ of disjoint indexes of
    letters of $\pi$ such that $\pi_{|I^1}$ and $\pi_{|I^2}$ are both
    order-isomorphic to $\sigma$.
    Let $G_\pi$ be the linear clique associated to $\pi$ with vertices
    $\VS{G_\pi} = \{i^1_j : j \in [n]\} \cup \{i^2_j : j \in [n]\}$.
    Define $\vv{\mathcal{M}} = \{(i^1_j, i^2_j) : j \in [n]\}$.
    It is easily seen that $\vv{\mathcal{M}}$ is an oriented perfect
    matching since $I^1 \cap I^2 = \emptyset$, and
    $I^1 \cup I^2 = \{1, 2,  \dots 2n\}$.
    It also avoids \InclusionLL,~\InclusionLR,~\InclusionRL and \InclusionRR.
    Indeed, suppose, aiming at a contradiction, that such a bad configuration exists
    for arcs $(i^1_j, i^2_j)$ and $(i^1_k, i^2_k)$ of $\vv{\mathcal{M}}$
    with $j < k$.
    \begin{itemize}
      \item
      \LabeledInclusionLL{$i^1_j$}{$i^1_k$}{$i^2_k$}{$i^2_j$}
      \item
      \LabeledInclusionLL{$i^1_j$}{$i^1_k$}{$i^2_k$}{$i^2_j$}
      \item
      \LabeledInclusionLL{$i^1_j$}{$i^1_k$}{$i^2_k$}{$i^2_j$}
      \item
      \LabeledInclusionLL{$i^1_j$}{$i^1_k$}{$i^2_k$}{$i^2_j$}
    \end{itemize}

    Indeed, if it is not the case then there would exist two edges
    $e = \{i^1_j, i^2_j\}$ and $e' = \{i^1_k, i^2_k\}$, $j < k$, in
    $\mathcal{M}$ such that $i^1_j < i^1_k < i^2_k < i^2_j$ or
    $i^1_j < i^2_k < i^1_k < i^2_j$. This is a contradiction since
    $i^2_k > i^2_j$ if $k > j$. Finally, for any two distinct edges
    $e = \{\i^1_j, i^2_j\}$ and $e' = \{i^1_k, i^2_k\}$ of $\mathcal{M}$,
    we have $\pi(i^1_j) < \pi(i^2_j)$ if and only if
    $\pi(i^1_k) < \pi(i^2_k)$ since we are comparing in both cases two
    elements (at position $j$ and $k$) in two patterns that are
    order-isomorphic to $\sigma$.
